Continuous estimates of heat emission at Mt. Ruapehu using the Unscented Kalman Smoother

### 2021-07-01 manuscript/manuscript.tex
add more sections; change synthetic test section and bibliography

%%% Load packages
%\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

@@

\title{Continuous estimates of heat emission at Mt. Ruapehu using
the unscented Kalman Smoother}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={y.behr@gns.cri.nz}   % email address
]{\inits{YB}\fnm{Yannik} \snm{Behr}}
\author[
   addressref={aff1},
   email={s.sherburn@gns.cri.nz}
]{\inits{SS}\fnm{Steve} \snm{Sherburn}}

@@

@@

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc_template/bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{zotero.bib}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

>>>>

%%% Load packages
%\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{graphicx}

@@

\title{Continuous estimates of heat emission at Mt. Ruapehu using
the Unscented Kalman Smoother}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={y.behr@gns.cri.nz}   % email address
]{\inits{YB}\fnm{Yannik} \snm{Behr}}
\author[
   addressref={aff1},
   email={s.sherburn@gns.cri.nz}
]{\inits{SS}\fnm{Steve} \snm{Sherburn}}
\author[
   addressref={aff1},
   email={t.hurst@gns.cri.nz}
]{\inits{TH}\fnm{Tony} \snm{Hurst}}

@@

We propose a probabilistic (Bayesian) inference approach which naturally
handles these difficulties and transparently propagates uncertainties of
observations and prior assumptions into posterior uncertainties of heat and
steam input. The accurate description of uncertainties is of particular
importance when publishing automatically generated estimates of heat and steam
input, as they could potentially be used in decision making or automatic
alerting of monitoring staff.

\begin{figure}
\includegraphics[width=\textwidth]{figures/overview.png}  
\caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Karen
	Britten, GNS Science). The red O marks the location of the lake outlet. The
	location of Mt. Ruapehu is shown as a red triangle on the insert map.}
  \label{overview}
\end{figure}

\section*{Probabilistic inference method}\label{Pim}

\subsection*{Physical model}\label{phm}

Following \citet{Hurst1991} and \citet{Stevenson1992} the mass and energy balance
can be written in terms of three coupled ordinary differential equations. The
rate of lake water mass change can be described as:
\begin{equation}\label{ode_M}
	\frac{dM}{dt} = M_s - M_o - M_e + M_i
\end{equation}

Where $M_s$ is the rate of water inflow through melt and precipitation, $M_o$
is the outflow rate through seepage and overflow, $M_e$ is the rate of mass
loss through evaporation and $M_i$ is the rate of steam input through the vent
system beneath the lake.

The thermal energy content of a mass flow can be described with the notion of
enthalpy, $H$, which is the ratio of energy and mass. Enthalpy can also be defined
on a relative scale as:
\begin{equation}
	H(T) = H(T_0) + \int_{T_0}^{T}c_p\tau d\tau
\end{equation}

where $T$ is temperature and $c_p$ is the specific heat. For all following
equations we take $H(T_0=0)$ as our reference enthalpy and $c_p=c_w=constant$,
with $c_w$ the specific heat of water, such that the enthalpy of lake water at
temperature $T$ can be written as $H(T) = c_w T$. Because Ruapehu Crater Lake
is at an altitude of ~2500 m and big parts of its catchment are covered by snow
and ice all year round we assume the temperature of incoming water to be at 0
$^{\circ}C$.

The rate of energy change of the lake water can thus be written as:
\begin{equation}\label{ode_E}
    \frac{dE}{dt}=\frac{d(c_wTM)}{dt}=c_wT\frac{dM}{dt} + c_wM\frac{dT}{dt} =
	-Q_e - c_wTM_o + Q_{eff} + Q_r
\end{equation}

where $T$ is the lake water temperature, $M$ is the lake water mass, $Q_r$ is
the rate of energy gain due to solar irradiation (short-wavelength radiation),
and $c_wTM_o$ is the rate of energy loss through outflow/seepage. $Q_e$ is
the rate of surface energy losses due to evaporation (forced and free
convection), sensible heat, and long-wavelength radiation (net loss). Several
empirical and theoretical equations exist to describe the surface losses and
both \citet{Stevenson1992} and \citet{Hurst2015} discuss them in detail. We
follow here mostly the choice of \citet{Hurst2015} as it includes the recent
development by \citet{Sartori2000} on forced (i.e. wind-induced) convection
(see Appendix \ref{A} for more details). $Q_{eff}$ is the effective rate of
energy gain from volcanic sources after heating incoming water:
\begin{equation}
	Q_{eff}=Q_i - M_s c_w T
\end{equation}
with $Q_i$ the rate of energy input from volcanic sources.

Following from Equation \ref{ode_E} the temperature change $\frac{dT}{dt}$ can then be written as:
\begin{equation}\label{ode_T}
	\frac{dT}{dt}=\frac{1}{c_wM}\left(-Q_e - c_wTM_o + Q_{eff} + Q_r\right )
	-\frac{T}{M}\frac{dM}{dt}
\end{equation}

Finally, the change in ion concentration of, for example, $Mg^{2+}$ can be expressed as:
\begin{equation}\label{ode_X}
	\frac{dX}{dt}=-M_o\frac{X}{M}
\end{equation}

Where $X$ is the total ion amount in the lake. This equation is only true under
the assumption that $X$ only changes due to dilution and there is no re-supply
between eruptive episodes. Equation \ref{ode_X} gives an estimate of the average
outflow over periods on the order of months but does not capture shorter term
variations in the outflow.

\subsection*{Inference}\label{inf}

The objective of the inference problem is to estimate the amount and energy
content of steam and gas entering the lake from the hydrothermal system below.
To constrain these we use the currently available observations of lake
temperature ($T$), lake level ($L$), ion concentration ($X$) as well as
numerical weather model data on windspeed. Putting it more concisely, we are
looking for

\begin{equation}\label{bsm}
    p(\bm{x}_k|\bm{y}_{1:T})
\end{equation}

where $\bm{y}_{1:T}$ is the time series of $m$ observations and $x_k$ comprises
the $n$ model parameters described in Section \ref{phm} at time step $k$. Note
that $x_k$ also includes the lake's temperature, lake level, and ion
concentration. We distinguish here between the state of the lake and
observations of this state ($T$, $L$, $X$) which are to some degree uncertain.

The following equations are based on \citet{Sarkka2010} which we refer the
reader to for more proofs and more detailed explanations.

Equation \ref{bsm} is the marginal probability of $\bm{x}_k$ given all
observations until time $T$. We assume that the states of the lake form a
Markov chain, meaning that the state of the lake at time step $k$, $\bm{x}_k$,
only depends on the state at adjacent time steps:

\begin{equation}\label{markov1}
    p(\bm{x}_k|\bm{x}_{1:k-1}, \bm{y}_{1:k-1}) = p(\bm{x}_k|\bm{x}_{k-1})
\end{equation}

and

\begin{equation}\label{markov2}
    p(\bm{x}_{k-1}|\bm{x}_{k:T}, \bm{y}_{k:T}) = p(\bm{x}_{k-1}|\bm{x}_{k})
\end{equation}

With this assumption and using \textit{Bayes' rule}, Equation \ref{bsm} can be
rewritten as:

\begin{equation}\label{bsm1}
p(\bm{x}_k|\bm{y}_{1:T}) = p(\bm{x}_k|\bm{y}_{1:k})
    \int\left[\frac{p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_{k+1}|\bm{y}_{1:T})}{p(\bm{x}_{k+1}|\bm{y}_{1:k})}\right]d\bm{x}_{k+1}
\end{equation}

with

\begin{eqnarray}\label{bfilt1}
p(\bm{x}_{k+1}|\bm{y}_{1:k}) & = & \int p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_k|\bm{y}_{1:k})d\bm{x}_k \\
        \label{bfilt}
    p(\bm{x}_k|\bm{y}_{1:k}) & = & \frac{1}{z_k} p(\bm{y}_k|\bm{x}_k)p(\bm{x}_k|\bm{y}_{1:k-1}) \\
        \label{zetk}
    z_k & = & \int p(\bm{y}_k|\bm{x}_k) p(\bm{x}_k| \bm{y}_{1:k-1}) d\bm{x}_k  
\end{eqnarray}

using the \textit{Chapman-Kolmogorov} equation we can write:

\begin{equation}
p(\bm{x}_k|\bm{y}_{1:k-1}) = \int p(\bm{x}_k|\bm{x}_{k-1})p(\bm{x}_{k-1}|\bm{y}_{1:k-1})d\bm{x}_{k-1} 
\end{equation}

Equations \ref{bsm1} to \ref{zetk} describe a recursive way to compute Equation
\ref{bsm} in which we only need to know $\bm{x}_1$,
$p(\bm{x}_k|\bm{x}_{k-1})$, and $p(\bm{y}_k|\bm{x}_k)$. Equation \ref{bsm} and
\ref{bfilt} are also known as the Bayesian Smoothing and Bayesian Filtering
equation, respectively. If $p(\bm{x}_k|\bm{x}_{k-1})$ and
$p(\bm{y}_k|\bm{x}_k)$ are linear transformations with normally distributed
errors, the \textit{Kalman filter} \citep{Kalman1960} and Kalman smoother
\citep{Rauch1965} represent closed form solutions. In our case, these
transformations are:

\begin{eqnarray}
p(\bm{x}_k|\bm{x}_{k-1}) & = & f(\bm{x}_{k-1}) + \bm{q}_{k-1} \\
    p(\bm{y}_k|\bm{x}_k) & = & \bm{x}_k[1:m] + \bm{r}_k   
\end{eqnarray}

where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
which is highly non-linear. If we assume the errors, $\bm{r}_k$ and
$\bm{q}_{k-1}$, to be normally distributed ($\bm{q}_{k-1} \sim N(\bm{0},
\bm{Q}_{k-1})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
approximate solution to \ref{bsm} using the \textit{Unscented Kalman Filter}
\citep{Merwe2004}.

We estimate $\bm{R}_k$ from the daily variations in the observations but we do
not know $\bm{x}_1$ and $\bm{Q}_{k-1}$. If $\bm{\theta}$ stands for all unknown
parameters than we are looking for $p(\bm{\theta}|\bm{y}_{1:T})$. Using again
\textit{Bayes' Theorem} and ignoring the normalising constant we can write:

\begin{equation}\label{pe1}
    p(\bm{\theta}|\bm{y}_{1:T}) \propto p(\bm{y}_{1:T}|\bm{\theta})p(\bm{\theta})
\end{equation}

Assuming that $\bm{Y}_{1:T}$ form a Markov chain and that $p(\bm{\theta})$ follows
a uniform distribution we can rewrite Equation \ref{pe1} as:

\begin{equation}\label{pe2}
    p(\bm{\theta}|\bm{y}_{1:T}) \propto \prod_{k=1}^Tp(\bm{y}_{k}|\bm{y}_{k-1}, \bm{\theta})
\end{equation}

Each factor on the right-hand side of Equation \ref{pe2} can be written as:
\begin{equation}
    p(\bm{y}_k|\bm{y}_{1:k-1}, \bm{\theta}) = \int p(\bm{y}_k|\bm{x}_k, \bm{\theta})p(\bm{x}_k|\bm{y}_{1:k-1})d\bm{x}_k
\end{equation}

which is readily computed as a by-product of Equation \ref{bfilt}. We then run
a non-linear optimisation to find the \textit{maximum a posteriori (MAP)}
estimate of Equation \ref{pe2}.

* sensitivity analysis: high vs. low lake temperature; windspeed dependence;
  enthalpy dependence

\section*{Results}
\subsection*{Synthetic test}\label{syn_test}
To test the inversion scheme we design a synthetic test, consisting of a very
large cylinder, the same volume as Ruapehu Crater Lake. We keep the windspeed
and surface water inflow constant at 4.5 m/s and 10 kt/day, respectively. To
model the outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section
and then use Bernoulli's equation to model the water outflow based on the water
level in the cylinder above the pipe. We assume that the incoming heat follows
a combination of a third-order polynomial and a sinusoid function and that the
enthalpy of the incoming steam is 2.8 which corresponds to the enthalpy of
saturated steam at hydrostatic pressures similar to those at the bottom of the
lake \ref{Mayhew1978}. To the resulting synthetic observations we add
uncertainties similar to those estimated from real observations (see Section
\ref{realdata}).

Figure \ref{syn_example} shows the result of the probabilistic inference of
heat input compared to the true input of the synthetic model. The observations
are well fit within their error bounds but we see some descrepancies between
the original and the recovered heat input rates, especially for very steep
changes in the input rate. This is the result of preferring a smooth solution
as already mentioned in \ref{inf}.

It is noteworthy that the mass influx and outflux rates are recovered quite accurately despite having no data in the former and very little data in the latter case.

\begin{figure}
	\includegraphics[width=.9\textwidth]{figures/synthetic_inversion_uks.png}  
    \caption{Inversion result for the synthetic test described in Section 
        \ref{syn_test}. Turquoise lines and markers represent hypothetical
        observations; red dashed lines are the inversion results and black
        dashed lines are the true input to the synthetic test.}
  \label{syn_example}
\end{figure}

@@

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{zotero}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}
